---
title: "Two-theta model analysis"
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ape)
library(ggiraph)

library(evemodel)
```

```{r loadData}
tree <- read.tree("data/comparison/newFormat/salmonData/salmonidsPhylo.newick")
gene.data <- getExprData("data/comparison/newFormat/salmonData/salmon20.tsv")

shiftSpecies <- c("Salp","Okis","Omyk","Ssal")
initParams <- initialParamsTwoTheta(gene.data,colSpecies = colnames(gene.data),shiftSpecies = shiftSpecies)

lowerBound <- c(theta1 = -99, theta2 = -99, sigma2 = 0.0001, alpha = 0.001, beta = 0.001)
upperBound <- c(theta1 =  99, theta2 =  99, sigma2 =   9999, alpha = 999  , beta = 99   )
```

```{r defFun}

parTransFuns <- list(
  none = list(
    trans = function(par){par},
    untrans = function(par){par}
  ),
  log = list(
    trans = function(par){
      with(as.list(par),
           c( theta1 = theta1, theta2 = theta2, 
              sigma2 = log(sigma2), alpha = log(alpha), beta = log(beta)))
    },
    untrans = function(par){
      with(as.list(par),
           c( theta1 = theta1, theta2 = theta2, 
              sigma2 = exp(sigma2), alpha = exp(alpha), beta = exp(beta)))
    }),
  rho = list(
    trans = function(par){
      with(as.list(par),
           c( theta1=theta1, theta2=theta2, alpha=alpha,
              rho=sigma2/(2*alpha), tau=beta*sigma2/(2*alpha)))
    },
    untrans = function(par){
      with(as.list(par),
           c( theta1=theta1, theta2=theta2, sigma2 = 2*rho*alpha,
              alpha=alpha, beta = tau/rho))
    })
)

fitTwoThetasTrans <- function( tree, shiftSpecies, colSpecies, gene.data.row, 
                               initParams, lowerBound, upperBound,
                               trans = c("none","log","rho"),
                               shiftOnStartOfBranch = F){
  
  transFun <- parTransFuns[[match.arg(trans)]]

  isThetaShiftEdge <- 1:Nedge(tree) %in% getEdgesFromMRCA(tree, tips = shiftSpecies)
  
  if( shiftOnStartOfBranch ) {
    isThetaShiftEdge[match(getMRCA(tree,shiftSpecies), tree$edge[ ,2])] <- TRUE
  }
  
  # match the column species with the phylogeny tip labels
  index.expand <- match(colSpecies, tree$tip.label)

  # function to optimize
  LLPerGeneTwoTheta <- function(par)
  {
    params <- transFun$untrans(par)
    ll <- logLikTwoTheta( theta1 = params[1], theta2 = params[2], sigma2 = params[3],
                          alpha = params[4], beta = params[5],
                          tree = tree, isThetaShiftEdge = isThetaShiftEdge, 
                          gene.data.row = gene.data.row, index.expand = index.expand)
    return(-ll)
  }
  
  # This doesn't seem to work..
  # if( trans == "rho"){
  #   # rho transform uses inverse of alpha, therefore swap upper and lower bound of alpha
  #   tmp <- lowerBound["alpha"]
  #   lowerBound["alpha"] <- upperBound["alpha"]
  #   upperBound["alpha"] <- tmp
  # }
    
  res <- optim( par = transFun$trans(initParams), fn = LLPerGeneTwoTheta, method = "L-BFGS-B",
                lower = transFun$trans(lowerBound), 
                upper = transFun$trans(upperBound))

  res$par <- transFun$untrans(res$par)
  
  return(res)
}

```

```{r fitAll, cache=T}
lapply(c("log","none") %>% setNames(.,.), function(trans){
  lapply(rownames(gene.data), function(geneID){
    fitTwoThetasTrans(tree = tree, shiftSpecies = shiftSpecies, colSpecies =colnames(gene.data), 
                      gene.data.row = gene.data[geneID, ], 
                      initParams = initParams[geneID, ], 
                      trans = trans,
                      lowerBound = lowerBound, upperBound = upperBound)
  })  
}) -> resAll

resTbl <- 
  lapply(resAll,function(resTrans){
    t(sapply(resTrans, function(res) res$par)) %>% 
    as.tibble() %>% 
    mutate( geneID=rownames(gene.data),
            ll = sapply(resTrans, function(res) -res$value),
            convergence = map_int(resTrans, "convergence"),
            iterations = sapply(resTrans, function(res) res$counts[1]))
  }) %>% 
  bind_rows(.id = "trans")
```

## log transform and bounds

### effect on iterations

```{r plotIterations}
resTbl %>% 
  mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
  mutate( convergence = convergence==0 ) %>% 
  ggplot( aes(x=geneID,y=iterations, fill=trans, color=convergence)) +
  geom_col(position = "dodge",width = 0.7) +
  scale_color_manual( values = c("green","black")) +
  coord_flip()
```

### parameter estimates

```{r plotPars}

resTbl %>%
  select(-theta1,-theta2, -iterations) %>% 
  group_by(geneID) %>%
  mutate( llDiff = ll - mean(ll)) %>% 
  ungroup() %>% 
  select(-ll) %>% 
  gather(key = "Parameter", value = "Value", -geneID, -trans, -convergence, -llDiff) %>%
  mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
  mutate( atUpperBound = Value > (upperBound[Parameter]*0.999)) %>%
  mutate( atLowerBound = Value < (lowerBound[Parameter]*1.001)) %>%
  mutate( atBounds = atUpperBound | atLowerBound) %>% 
  ggplot( aes(x=Value,y=geneID, color=llDiff, shape=trans)) + 
  geom_point( size=3) + 
  scale_shape_manual(values = c(3,4)) +
  scale_color_gradient2(low="blue",mid="black", high="red") +
  facet_grid(. ~ Parameter ,scales = "free") +
  theme(legend.position="bottom") +
  scale_x_log10() +
  geom_vline(data=tibble(x=upperBound[3:5],Parameter=names(upperBound[3:5])),aes(xintercept=x),
             color="red",linetype=3)

```


## overfitting alpha to the mean?

### Expected expression {.tabset}
```{r plotExp,results="asis"}
# plot the expression on Y and time since shift on the X. + Expected expression trend line.

# get time since shift
mrcaNode <- getMRCA(tree,shiftSpecies)
spcNodes <- match(shiftSpecies,tree$tip.label)
timeSinceShift <- setNames(dist.nodes(tree)[mrcaNode,spcNodes], shiftSpecies)

for( gene in rownames(gene.data)){

  cat("\n\n####",gene,"\n")
  
  with(as.list(resTbl %>% filter(geneID==gene, trans=="log")),{
    trendLine <- 
      tibble(timeSinceShift = seq(0,max(timeSinceShift),length.out = 100)) %>% 
      mutate( expLevel = theta2 + (theta1-theta2) * exp(-alpha * timeSinceShift))
    
    tibble(expLevel=gene.data[gene,], spc=colnames(gene.data)) %>% 
      mutate( timeSinceShift = ifelse(spc %in% shiftSpecies,timeSinceShift[spc],0) ) %>% 
      ggplot( aes( x = timeSinceShift, y = expLevel)) +
      geom_point() +
      geom_line( data=trendLine, color="red",linetype=2) +
      geom_hline(yintercept=theta1, color="blue",linetype=3) +
      annotate( "text", label="theta1",x=0.02,y=theta1, color="blue", vjust=-0.5) +
      geom_hline(yintercept=theta2, color="blue",linetype=3) + 
      annotate( "text", label="theta2",x=0.02,y=theta2, color="blue", vjust=-0.5) +
      ggtitle(paste0(gene,"  sigma2 = ",signif(sigma2,3),"  alpha = ",signif(alpha,3), "  beta = ", signif(beta,3)))
  }) -> g
  print(g)
  
}

```

### Alpha and Ssal expression

Beacause of the phylogeny where Ssal has a significantly shorter branch than the other species and the decision to define the theta shift at the most recent common ancestor of the salmonids

```{r relSsalShiftPlot}
colSpeciesIndices <- split(1:ncol(gene.data), f = colnames(gene.data))
species.mean <- sapply(colSpeciesIndices, function(i){ rowMeans(gene.data[,i]) })

tibble(
  geneID = rownames(gene.data),
  preShift = rowMeans(species.mean[ ,c("Drer", "Eluc", "Olat")]),
  postShift = rowMeans(species.mean[ ,c("Salp", "Okis", "Omyk")]),
  Ssal = species.mean[ ,"Ssal"]
) %>% 
  mutate( relSsalShift = (Ssal-preShift) / (postShift-preShift)) %>% 
  left_join( resTbl %>% filter(trans=="log") %>% select(geneID,alpha), by="geneID") %>% 
  ggplot( aes(x=relSsalShift, y=alpha, color=alpha)) +
  scale_color_gradientn(colours = c("blue","black","red"),trans = "log", breaks = c(10^(-1:3))) +
  geom_point_interactive( aes(tooltip=geneID, data_id = geneID)) + 
  scale_y_log10() -> g

girafe(code = print(g))
```

### Between-/Within-species variance
```{r betweenSpeciesVarPlot}

# calculate expected mean based on parameters
expLevel <- species.mean*0
for( gene in rownames(species.mean)){
  with(as.list(resTbl %>% filter(geneID==gene, trans=="log")),{
    for( spc in colnames(species.mean)){
      dt <- ifelse( spc %in% names(timeSinceShift),timeSinceShift,0)
      expLevel[gene,spc] <<- theta2 + (theta1-theta2) * exp(-alpha * dt)
    }
  })
}

# subtract the expected mean before calculating between species mean
species.mean.adjusted <- species.mean - expLevel

tibble(
  geneID = rownames(gene.data),
  betweenSpeciesVar = apply(species.mean.adjusted,1,var)
) %>% 
  left_join( resTbl %>% filter(trans=="log"), by="geneID") %>% 
  ggplot( aes(x=betweenSpeciesVar, y=sigma2/(2*alpha), color=alpha)) +
  scale_color_gradientn(colours = c("blue","black","red"),trans = "log", breaks = c(10^(-1:3))) +
  geom_point() +
  geom_point_interactive( aes(tooltip=geneID, data_id = geneID)) + 
  geom_abline( slope=1, intercept=0, color="red", linetype=2) +
  scale_x_log10() +
  scale_y_log10() -> g

girafe(code = print(g))  
```

```{r withinSpeciesVarPlot}
tibble(
  geneID = rownames(gene.data),
  withinSpeciesVar = rowMeans(sapply(colSpeciesIndices, function(i){ apply(gene.data[,i],1,var) }))
) %>% 
  left_join( resTbl %>% filter(trans=="log"), by="geneID") %>% 
  ggplot( aes(x=withinSpeciesVar, y=beta*sigma2/(2*alpha), color=alpha)) +
  scale_color_gradientn(colours = c("blue","black","red"),trans = "log", breaks = c(10^(-1:3))) +
  geom_point() +
  geom_point_interactive( aes(tooltip=geneID, data_id = geneID)) + 
  geom_abline( slope=1, intercept=0, color="red", linetype=2) +
  scale_x_log10() +
  scale_y_log10() -> g

girafe(code = print(g))  
```

## Late vs early theta-shift

In the model used above the theta-shift was placed at the point of the first diverging species in the salmonid clade. However, the WGD event, which is were we expect the shift, occurred earlier at some point along the branch leading to the salmonids. As we have seen above, the relatively short Ssal branch leads to overfitting to the Ssal expression. By moving the theta-shift to the start of the salmonid branch there is less difference in the time since shift and there should be less overfitting.

```{r plotTreeWithShiftColors}
orangeEdges <- 1:Nedge(tree) %in% getEdgesFromMRCA(tree, tips = shiftSpecies)

plot.phylo(tree, edge.color = ifelse(orangeEdges,"orange","black"),
           tip.color = ifelse(tree$tip.label %in% shiftSpecies,"orange","black"),
           main = "Shift at root of clade")
  
orangeEdges[match(getMRCA(tree,shiftSpecies), tree$edge[ ,2])] <- TRUE

plot.phylo(tree, edge.color = ifelse(orangeEdges,"orange","black"),
           tip.color = ifelse(tree$tip.label %in% shiftSpecies,"orange","black"),
           main = "Shift at start of branch to clade")
```

```{r fitEarlyShift, cache=TRUE}
lapply(rownames(gene.data), function(geneID){
    fitTwoThetasTrans(tree = tree, shiftSpecies = shiftSpecies, 
                      colSpecies =colnames(gene.data), 
                      gene.data.row = gene.data[geneID, ], 
                      initParams = initParams[geneID, ], 
                      trans = "log",
                      lowerBound = lowerBound, upperBound = upperBound,
                      shiftOnStartOfBranch = TRUE)
}) -> resEarlyShift


resTblEarly <- 
  t(sapply(resEarlyShift, function(res) res$par)) %>% 
  as.tibble() %>% 
  mutate( geneID=rownames(gene.data),
          ll = sapply(resEarlyShift, function(res) -res$value),
          convergence = map_int(resEarlyShift, "convergence"),
          iterations = sapply(resEarlyShift, function(res) res$counts[1]))

```

### Compare estimated parameters

```{r plotCompareParamsEarlyLate}

bind_rows(.id = "shiftPoint", 
          early = resTblEarly,
          late = filter(resTbl, trans =="log")) %>% 
  select(-theta1,-theta2, -iterations, -trans, -convergence) %>% 
  group_by(geneID) %>% 
  mutate( llDiff = ll - mean(ll)) %>% 
  ungroup() %>% 
  select(-ll) %>% 
  gather(key = "Parameter", value = "Value", sigma2, alpha, beta) %>%
  mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
  ggplot( aes(x=Value,y=geneID, color=llDiff, shape=shiftPoint)) + 
  geom_point( size=3) + 
  scale_shape_manual(values = c(3,4)) +
  scale_color_gradient2(low="blue",mid="black", high="red") +
  facet_grid(. ~ Parameter ,scales = "free") +
  theme(legend.position="bottom") +
  scale_x_log10() +
  geom_vline(data=tibble(x=upperBound[3:5],Parameter=names(upperBound[3:5])),aes(xintercept=x),
             color="red",linetype=3)

```

```{r plotCompareLL}
resTblEarly %>% 
  transmute(llEarly=ll,geneID=geneID) %>% 
  left_join(resTbl %>% filter( trans =="log") %>% select(ll,geneID), by="geneID") %>% 
  mutate( llDiff = llEarly - ll) %>% 
  mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
  ggplot( aes(y=llDiff, x=geneID, fill=llDiff >0)) + 
  geom_col() +
  scale_fill_manual(values = c("blue","red"), guide=F) +
  coord_flip() + 
  ylab("logL_early - logL_late")
```

The likelihoods, with few exceptions, are lower when using the early shift. Is this because the late shift model reflects the true process better or is it because the late shift phylogeny allows more overfitting?

### Expected expression {.tabset}

```{r plotExpEarlyShift,results="asis"}
# plot the expression on Y and time since shift on the X. + Expected expression trend line.

# get time since shift
earlyNode <- getMRCA(tree,c(shiftSpecies,"Eluc"))
spcNodes <- match(shiftSpecies,tree$tip.label)
timeSinceShift <- setNames(dist.nodes(tree)[earlyNode,spcNodes], shiftSpecies)

for( gene in rownames(gene.data)){

  cat("\n\n####",gene,"\n")
  
  with(as.list(resTblEarly %>% filter(geneID==gene)),{
    trendLine <- 
      tibble(timeSinceShift = seq(0,max(timeSinceShift),length.out = 100)) %>% 
      mutate( expLevel = theta2 + (theta1-theta2) * exp(-alpha * timeSinceShift))
    
    tibble(expLevel=gene.data[gene,], spc=colnames(gene.data)) %>% 
      mutate( timeSinceShift = ifelse(spc %in% shiftSpecies,timeSinceShift[spc],0) ) %>% 
      ggplot( aes( x = timeSinceShift, y = expLevel)) +
      geom_point() +
      geom_line( data=trendLine, color="red",linetype=2) +
      geom_hline(yintercept=theta1, color="blue",linetype=3) +
      annotate( "text", label="theta1",x=0.02,y=theta1, color="blue", vjust=-0.5) +
      geom_hline(yintercept=theta2, color="blue",linetype=3) + 
      annotate( "text", label="theta2",x=0.02,y=theta2, color="blue", vjust=-0.5) +
      ggtitle(paste0(gene,"  sigma2 = ",signif(sigma2,3),"  alpha = ",signif(alpha,3), "  beta = ", signif(beta,3)))
  }) -> g
  print(g)
  
}

```

### Alpha vs Ssal relative expression

```{r relSsalEarlyShiftPlot}
colSpeciesIndices <- split(1:ncol(gene.data), f = colnames(gene.data))
species.mean <- sapply(colSpeciesIndices, function(i){ rowMeans(gene.data[,i]) })

tibble(
  geneID = rownames(gene.data),
  preShift = rowMeans(species.mean[ ,c("Drer", "Eluc", "Olat")]),
  postShift = rowMeans(species.mean[ ,c("Salp", "Okis", "Omyk")]),
  Ssal = species.mean[ ,"Ssal"]
) %>% 
  mutate( relSsalShift = (Ssal-preShift) / (postShift-preShift)) %>% 
  left_join( resTblEarly %>% select(geneID,alpha), by="geneID") %>% 
  ggplot( aes(x=relSsalShift, y=alpha, color=alpha)) +
  scale_color_gradientn(colours = c("blue","black","red"),trans = "log", breaks = c(10^(-1:3))) +
  geom_point_interactive( aes(tooltip=geneID, data_id = geneID)) + 
  scale_y_log10() -> g

girafe(code = print(g))
```

With the early theta-shift the 

